---
title: "CLT Convergence: Six-Sided Die"
format:
  html:
    toc: true
    code-fold: show
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
source("../R/clt_functions.R")

student_id <- 8515     # last 4 digits of my student ID
B          <- 10000    # number of simulated samples (NOT the sample size)
pop        <- "die"
```

## 1. What does the population look like?

X is uniform on {1, 2, 3, 4, 5, 6}, generated with `sample()`. Discrete, symmetric, flat.

```{r population}
plot_population(pop, seed = student_id)
```

The population is discrete: each value from 1 to 6 has equal probability 1/6, so the bar chart is flat. It is perfectly symmetric (skewness = 0) but has no tails at all: values can never go below 1 or above 6. It is flatter than a normal curve, with excess kurtosis of about −1.27. It satisfies every CLT assumption (independent, identically distributed, finite variance), so the CLT applies. The main features to watch are the flat shape and the gaps between possible values.

## 2. Theoretical mean and standard deviation

Theoretical mean = 3.5, SD = √(35/12) ≈ 1.708, so SE(X̄) = 1.708/√n.

## 3. Prediction (write this BEFORE running the simulations)

Prediction: The die is symmetric, so I expect skewness to be close to 0 at every n. Because it is flat (uniform) with no tails, it is already closer to normal than the exponential, so I predict it will look normal fairly quickly, around n = 5–10, well under 30. Since the die is discrete, I expect the histogram to look choppy at small n. I expect the simulated mean to be near 3.5 and the SE to match 1.708/√n.

## 4. Which sample sizes did I investigate?

**Pass 1 (coarse):** a wide grid to find roughly where the change happens.

```{r pass1}
ns_coarse <- c(2, 5, 10, 20, 30, 50, 100, 200, 500)
coarse    <- run_study(pop, ns_coarse, B = B, seed = student_id)
knitr::kable(coarse$table, digits = 4)
plot_convergence(coarse$table, pop)
```

**Pass 2 (refined):** narrower values around the transition.

```{r pass2}
ns_fine <- c(3, 4, 5, 6, 7, 8, 9, 10, 12, 15)   # pass 1 changed between n = 5 and n = 10
fine    <- run_study(pop, ns_fine, B = B, seed = student_id)
knitr::kable(fine$table, digits = 4)
```

In pass 1, n = 5 failed my standard and n = 10 passed, so the change happens between them. I refined with every value from 3 to 10, plus 12 and 15, to find the exact crossing point.

## 5. Simulated vs. theoretical mean and standard error

The table columns `sim_mean` / `theo_mean` and `sim_se` / `theo_se` compare the simulation with the CLT's predictions. `se_ratio` = simulated SE ÷ σ/√n, so 1 means a perfect match.

The simulated mean was within 0.004 of the theoretical mean of 3.5 at every n, and the SE ratio stayed between 0.98 and 1.00. The simulation agrees with σ/√n = 1.708/√n at every sample size, even at n = 2 where the shape is not normal.

## 6. My "normal enough" standard and the chosen n

```{r criteria}
# My "normal enough" standard, used identically in every file
criteria <- list(
  max_abs_skew = 0.25,    # mild skew; tighter is within Monte Carlo noise
  max_abs_kurt = 0.3,     # tails close to normal
  min_qq_cor   = 0.998    # Q-Q plot essentially straight
)

all_results <- rbind(coarse$table, fine$table)
all_results <- all_results[!duplicated(all_results$n), ]
all_results <- all_results[order(all_results$n), ]
all_results$passes <- passes_standard(all_results, criteria)
knitr::kable(all_results, digits = 4)

smallest_normal_n(all_results, criteria)
```

```{r plots, fig.width = 9, fig.height = 3.5}
# Show values below, at, and above my chosen n
for (n in c(2, 9, 30)) {  
  plot_diagnostics(simulate_means(populations[[pop]], n, B, student_id), n, pop)
}
```

$$\boxed{n = \text{9}}$$

I used the same standard as every other population: |skewness| ≤ 0.25, |excess kurtosis| ≤ 0.3, and Q-Q correlation ≥ 0.998. Skewness passed at every n. Kurtosis passed from n = 5 on (−0.24). The Q-Q correlation was the last to pass: it was 0.99795 at n = 8, just below the cutoff, and 0.99813 at n = 9, and it stayed above 0.998 at every larger n. The Q-Q plot at n = 2 shows clear steps, while at n = 9 the points follow the line closely with only small steps. The histogram at n = 9 is bell-shaped and matches the red curve. The Shapiro-Wilk test still rejected normality until about n = 100, because it detects the discreteness that remains, which does not matter in practice.

## 7. Limitations and surprises

I expected skewness to be the main issue, as it was for the exponential, but the die is symmetric, so skewness stayed near 0 (|skewness| ≤ 0.04) at every n. Instead, two other features held it back. First, the die is flatter than a normal curve, with light tails and no extreme values, so the excess kurtosis started negative (−0.65 at n = 2) and only passed my cutoff at n = 5. This matches theory: the excess kurtosis of the mean is −1.27/n. Second, the die is discrete. At n = 2 the average can only take 11 different values, so the histogram has gaps and the Q-Q plot looks like a staircase. The number of possible averages grows with n (26 at n = 5, 38 at n = 10), and the staircase smooths out. Discreteness, not skew, was the last thing to disappear.
